### 2013-02-14 doc/xps2013/proposal.tex
Incorporated Preston's suggestions 	modified:   proposal.bib 	modified:   proposal.tex
fl

% Most importantly, however, for a much lower real world cost,
% users can \emph{add} significantly more processors to a commodity cluster than
% an XMT machine and use Grappa to achieve scalable performance.

@@

@@

@@

\paragraph{Global memory addressing} Grappa provides two methods for storing data in the global memory. The first is a distributed heap striped across all the machines in the system in a block cyclic fashion. The \texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate and deallocate memory in the global heap.  Addresses to memory in the global heap use \textbf{linear addresses}.  Choosing the block size involves trading off sequential bandwidth against aggregate random access bandwidth. Smaller block sizes help spread data across all the memory controllers in the cluster, but larger block sizes allow the locality-optimized memory controllers to provide increased sequential bandwidth. The block size, which is configurable, is typically set to 64 bytes, or the size of a single hardware cache line, in order to exploit spatial locality when available. The heap metadata is stored on a single node. Currently all heap operations serialize through this node; while this has been sufficient for our benchmarks, in the future Grappa will provide parallel performance through combining~\cite{MAMA,flatcombining}.

Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system. Addresses to global memory allocated in this way use \textbf{2D global addresses}.  This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process. The lower 48 bits of the address hold a virtual address in the process. The top bit is set to indicates that the reference is a 2D address (as opposed to linear address). This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints. Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address. This address can then be accessed with a delegate.
% and can also be cached by other nodes. 
At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address. 2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.

@@

% Delegate operations are always executed at the home core of their address, and
% while arbitrary memory operations can be delegated, we restrict the use of
% delegate operations in three ways to make them more useful for
% synchronization. First, we limit each task to one outstanding delegate
% operation to avoid the possibility of reordering in the network. Second, we
% limit delegate operations to operate on objects in the 2D address space or
% objects that fit in a single block of the linear address space, so they can be
% satisfied with a single network request. Finally, no context switches are
% allowed while the data is being modified. Given these restrictions, we can
% ensure that delegate operations for the same address from multiple requesters
% are always serialized through a single core in the system, providing atomic
% semantics without using atomic operations.

@@

% The network layer is serviced by polling. Periodically when a context switch
% occurs, the Grappa scheduler switches to the network polling thread. This
% thread has three responsibilities. First, it polls the lower-level network
% layer to ensure it makes progress. Second, it deaggregates received messages
% and executes active message handlers. Third, it checks to see if any
% aggregation queues have messages that have been waiting longer than the
% threshold; if so, it sends them.

@@

Some networks provide access to a remote machine's memory
directly. This would seem to be a good fit for a programming model
focused on global shared memory, but in fact we do not use it. In our
experiments, we found that RDMA operations are subject to the same
message rate limitations as all other messages on these cards, and
thus using raw RDMA operations for our small messages would make
inefficient use of bandwidth. Namely, we would be limited to the
network injection rate providing just a few million accesses per
second per node.  Instead, we implement remote memory operations with
active messages in conjuction with aggregation, providing an order of
magnitude or more increase in throughput. A byproduct of this design
decision is that Grappa is not limited to RDMA-capable networks.

@@

PI Oskin has traditionally participated in the computer-architecture research
community, co-PI Ceze, in addition to the architecture community also has a
presence in the systems and programming languages community, and co-PI Kahan
has participated mainly in the parallel and scientific computing community.
The move toward multicore architectures has re-invigorated collaboration
across these divides and recent joint work by Oskin, Ceze, and Kahan
embodies this important trend. There are currently five graduate students for
which last two of the PIs serve as advisors. In addition to the project
described in this proposal, they are working together on techniques for
detecting concurrency errors with machine learning and language
specifications, using disciplined approximations to save energy in computing,
and developing new operating-system abstractions for emerging non-volatile
memory devices. What has made these collaborations so fruitful in such a short
period of time is the PIs' complementary backgrounds. We now briefly describe
these backgrounds.

>>>>

@@

%%%%%%%%%%%%%% Tempting to delete from here....

@@

%%%%%%%%%%%%%%%%%  To here.... -Mark

@@

\paragraph{Global memory addressing} Grappa provides two methods for storing data in the global memory. The first is a distributed heap striped across all the machines in the system in a block cyclic fashion.  Addresses to memory in the global heap use \textbf{linear addresses}.  Choosing the block size involves trading off sequential bandwidth against aggregate random access bandwidth. Smaller block sizes help spread data across all the memory controllers in the cluster, but larger block sizes allow the locality-optimized memory controllers to provide increased sequential bandwidth. The block size, which is configurable, is typically set to 64 bytes, or the size of a single hardware cache line, in order to exploit spatial locality when available. The heap metadata is stored on a single node. Currently all heap operations serialize through this node; while this has been sufficient for our benchmarks, in the future Grappa will provide parallel performance through combining~\cite{MAMA,flatcombining}.

Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system. Addresses to global memory allocated in this way use \textbf{2D global addresses}.  This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process. The lower 48 bits of the address hold a virtual address in the process. The top bit is set to indicates that the reference is a 2D address (as opposed to linear address). This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints. Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address. This address can then be accessed with a delegate.  At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address. 2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.

@@

@@

@@

Some networks provide access to a remote machine's memory directly. This would seem to be a good fit for a programming model focused on global shared memory, however, we do not currently use it. In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations does not measurably increase achieved bandwidth (i.e., we are limited by injection rate not peak bandwidth).  At the moment, we implement remote memory operations with active messages in conjunction with aggregation, providing an order of magnitude or more increase in throughput compared to straightforward RDMA. As we describe later, however, we wish to explore richer semantics for RDMA communication that will make it integrate better with user level runtime systems such as Grappa.

@@

PI Oskin has traditionally participated in the computer-architecture research
community, co-PI Ceze, in addition to the architecture community also has a
presence in the systems and programming languages community, and co-PI Kahan
has participated mainly in the parallel and scientific computing community.
The move toward multicore architectures has re-invigorated collaboration
across these divides and recent joint work by Oskin, Ceze, and Kahan
embodies this important trend. There are currently five graduate students for
which the last two of the PIs serve as advisors. In addition to the project
described in this proposal, they are working together on techniques for
detecting concurrency errors with machine learning and language
specifications, using disciplined approximations to save energy in computing,
and developing new operating-system abstractions for emerging non-volatile
memory devices. What has made these collaborations so fruitful in such a short
period of time is the PIs' complementary backgrounds. We now briefly describe
these backgrounds.
